---
title: "R Notebook"
output: github_document
---

# Final Report Project 2


Reading the data and loading libraries.

```{r message=FALSE, warning=FALSE}
library(DBI)
library(RSQLite)
library(dplyr)
library(tidyr)
library(ggplot2)
library(janitor)
library(purrr)
library(tibble)
library(stringr)
library(infer)
library(lubridate)
library(leaflet)
library(httr)
library(magrittr)
library(purrr)

politicians <- dbConnect(RSQLite::SQLite(), "zh_politicians.db")

dbListTables(politicians)

mandates_tbl<-tbl(politicians, "MANDATES") %>% 
  collect()
addresses_tbl<-tbl(politicians, "ADDRESSES") %>% 
  collect()
affiliations_tbl<-tbl(politicians, "AFFILIATIONS") %>% 
  collect()
persons_tbl<-tbl(politicians, "PERSONS") %>% 
  collect()

dbDisconnect(politicians)
```

## Part 1

The MANDATES table has information about when people were elected, when their mandate ended and on which assembly they sat.

Using a line plot, show how the number of people with an active mandate changed over the years. You will have one line per assembly.

First some data cleaning. One observation has start year set at 0. The name of the person with start year 0 (ID 15150) is Johann Heinrich Furrer. He also has no birthdate. I googled his name and Grand Conseil and no results came up so I'll drop this observation. 
There are 183 rows with mandate end year 0. Seems that they probably all still hold the mandate as the start date is sufficiently recent. So I'll set these values to 2019.



```{r}
mandates_tbl <- mandates_tbl %>% 
  clean_names() %>% 
  mutate(mandate_end_year=as.numeric(mandate_end_year)) %>% 
  mutate(mandate_start_year=as.numeric(mandate_start_year))

mandates_tbl %>% 
  filter(mandate_start_year==0) 

persons_tbl %>% 
  filter(ID==15150)

mandates_tbl %>% 
  filter(mandate_end_year==0)

mandates_tbl <- mandates_tbl %>% 
  filter(mandate_start_year!=0) %>%  
  mutate(mandate_end_year=replace(mandate_end_year, mandate_end_year==0, "2019"))

```

Then I create a nested variable "years" listing all the years of active mandate. Then I unnest years.

```{r}
long_mand<- mandates_tbl %>% 
  mutate(years=map2(mandate_start_year, mandate_end_year, seq)) %>%   select(person_id, assembly, years) %>% 
  unnest(years)
  
long_mand %>% 
  count(assembly, years, name="people") %>% 
  ggplot(mapping = aes(x=years, y=people, color=assembly))+
  geom_line(size=1)+
  labs(title = "Number of active mandates each year (1804-2019)",
       subtitle = "The 'peaks' are caused by election years when multiple mandates were active for the same seat",
       y="n")+
  theme_light()
```


Don't bother working on exact dates: you can base your "active" years just on years, even if this creates some "dents" in the line (i.e. when looking only at active years, not precise dates, some mandates are double counted on election years since the old mandates end and the new mandates start).


## Part 2

Expand on the plot you just produced. This time you want to show facet charts with one chart per assembly. In each chart, have one line for men and one line for women.

To do this I first need to find a dataset including sex-disaggregated information. This is in the "persons" table. I first do some data cleaning. A count of "gender" shows that there are 4040 men, 249 women and 5 persons with no data. I filter the persons_tbl to have a look at the 5 persons. They can all be replaced with "m" given their names and birthdates/start of mandate. I then rename id with person_id for clarity.

```{r}
persons_tbl %>% 
  count(GENDER)

persons_tbl %>% 
  filter(GENDER=="") 

persons_tbl <- persons_tbl %>% 
  clean_names() %>% 
  mutate(person_id=id) %>%
  mutate(gender= replace(gender, gender=="", "m")) %>% 
  select(person_id, everything(), -id)

```

Then I join the two datasets and I unnest the years. This operation was slow on my computer so I keep it separate for the rest.
I then plot the chart with "free" scales as the smaller councils have much fewer mandates than the Cantonal Council.

```{r}
data2 <- left_join(mandates_tbl, persons_tbl, by=c("person_id")) %>% 
  select(person_id, assembly, mandate_start_year, mandate_end_year, lastname, firstname, gender) %>% 
  mutate(years=map2(mandate_start_year, mandate_end_year, seq)) %>%   unnest(years)

data2%>% 
  mutate(gender= replace(gender, is.na(gender), "m"),
         gender=factor(gender, levels=c("w", "m"))) %>% 
  select(person_id, gender, assembly, years) %>% 
  count(assembly, years, gender, name="people") %>% 
  ggplot(mapping = aes(x=years, y=people, color=gender))+
  geom_line(size=1)+
  facet_wrap(vars(assembly), scales="free")+
  labs(title = "Number of active mandates each year by sex",
       subtitle = "The 'peaks' are caused by election years when multiple mandates were active for the same seat",
       y="n")+
  theme_light()
```



## Part 3

Create a new plot showing the proportion of elected politicians from each party in year 2000. You want to show this by assembly, so use one facet with one pie chart per assembly. Also show your result in a table.

For this I need to find data on party affiliation. This in the affiliations file. I will do some data cleaning and then join this dataset with mandates.

```{r}
affiliations_tbl <- affiliations_tbl %>% 
  clean_names()

data3 <-
  left_join(mandates_tbl, affiliations_tbl, by=c("person_id")) %>% 
  select(mandate_id, person_id, party, assembly, mandate_start_year, mandate_end_year, affiliation_start_year, affiliation_end_year) %>% 
  filter(mandate_start_year<=2000) %>% 
  filter(mandate_end_year>=2000) %>% 
  mutate(party=str_replace(party, "NA", "others")) %>%
  mutate(party=if_else(party=="", "others", party)) %>% 
  count(assembly, party, name="number_of_elected_politicians") %>% 
  arrange(assembly, desc(number_of_elected_politicians)) %>%    mutate(party=factor(party, levels = c("SVP", "SP", "FDP", "CVP", "EVP", "GP", "Grüne", "SD", "SaS", "parteilos", "LdU", "POCH", "FPS", "EDU", "AL", "others")))

data3 %>% 
  group_by(assembly) %>% 
  mutate(tot_by_ass=sum(number_of_elected_politicians)) %>% 
  ungroup() %>% 
  mutate(share=number_of_elected_politicians/tot_by_ass) %>%
  mutate(lab=str_c(party, number_of_elected_politicians, sep = ", ")) %>% 
  ggplot(aes(x="", y=share, fill=party, label=lab))+
  geom_col(color="black")+
    coord_polar("y")+
  geom_text(position = position_stack(vjust = 0.5), check_overlap = TRUE)+
  facet_wrap(vars(assembly)) + 
  theme_void()+
  theme(legend.position="bottom", legend.direction="horizontal")+
  labs(title = "Elected officials by party affiliation in 2000")
```

## Part 4

Have another look at the composition of the assemblies: this time use a line chart to show how it changed over the years.

```{r fig.height=6}

data4 <- left_join(mandates_tbl, affiliations_tbl, by=c("person_id")) %>% 
  select(mandate_id, person_id, party, assembly, mandate_start_year, mandate_end_year, affiliation_start_year, affiliation_end_year) %>% 
  mutate(years=map2(mandate_start_year, mandate_end_year, seq)) %>%   unnest(years) 

list10 <- data4 %>% 
  select(mandate_id, person_id, party, assembly, years) %>% 
  mutate(party=str_replace(party, "NA", "other")) %>%
  mutate(party=if_else(party=="", "other", party)) %>%
  count(assembly, party, years, name="people") %>% 
  filter(years>=2015) %>% 
  count(party, name="num") %>% 
  arrange(desc(num)) %>% 
  head(10) %>% 
  pull(party) # creating a list of the top 10 parties by number of mandates during the entire period. This includes an unidentified party name. Thus the list is of 9 parties and "other".


data4 %>% 
  mutate(party10=if_else(party %in% list10, party, "other")) %>%
  mutate(party10=replace(party10, party10=="", "other")) %>% 
  select(mandate_id, person_id, party10, assembly, years) %>% 
  count(assembly, party10, years, name="people") %>%
  filter(assembly=="Cantonal Council" | assembly=="Executive Council") %>% 
  filter(years>=2000) %>% 
  ggplot(mapping = aes(x=years, y=people, color=party10))+
  geom_line()+
  facet_wrap(vars(assembly), ncol = 1, scales="free")+
  theme(legend.position="bottom", legend.direction="horizontal")+
  labs(title = "Elected officials by party affiliation 2000-2019",
       subtitle = "for the first ten parties by total number of mandates since 2015", 
       x="", y="")

```

I focused my chart on data from 2000 onwards and only for the Cantonal and the Executive Council. Prior to 1925 data by party affiliation is not very good. Most delegates end up in the "other" category because of missing data. Similarly party affiliation is not known for the Grand and Small Council mandates. 10 parties are included (all others become "other"), those with most total mandates for the 2015-2019 period.

## Part 5

For the politicians that have a YEAR_OF_DEATH in the data, find the average life span. Does it change if you go by a TITLE? Use first a plot, and then a statistical test to assess if the average life span is different for politicians having a TITLE compared to those without any TITLE.

We only ask you to compare TITLE vs. no TITLE, there is no need to analyse every possible title.

I first do some data cleaning. First I filter out all those with missing year of death. Then I set year of death as a numeric variable. Then I remove the one case when the year of birth was 0 as this is not possible and create a life span variable, whic I further use to remove some outliers in the data such as three people with life spans of 2,3 and 175 years, none of whih are humanly possible, so must be due to data entry typos. Finally I create a variable tnt (title, no title) which takes on 1 if the person has a title and 0 if the person has no title.


```{r}

ave_lfspan <- persons_tbl %>% 
  filter(year_of_death!="") %>%
  mutate(year_of_death=as.numeric(year_of_death)) %>% 
  filter(year_of_birth!=0) %>% 
  mutate(lfspan=year_of_death-year_of_birth) %>% 
  filter(lfspan>=4,  lfspan<=101) %>%
  summarise(ave_lfspan=mean(lfspan)) %>% 
  pull(ave_lfspan)

persons_tbl %>% 
  filter(year_of_death!="") %>%
  mutate(year_of_death=as.numeric(year_of_death)) %>% 
  filter(year_of_birth!=0) %>% 
  mutate(lfspan=year_of_death-year_of_birth) %>% 
  filter(lfspan>=4,  lfspan<=101) %>%
  mutate(tnt=if_else(title!="", "1", "0")) %>% 
  group_by(tnt) %>% 
  summarise(ave_lfspan=mean(lfspan))

```

The overall average lifespan is `r ave_lfspan`.Those without a title have a mean lifespan of 68 and those with a title of 74. There seems to be a difference, but is it statistically significant?

First a boxplot of the difference in distributions below. 


```{r}
persons_tbl %>% 
  filter(year_of_death!="") %>%
  mutate(year_of_death=as.numeric(year_of_death)) %>% 
  filter(year_of_birth!=0) %>% 
  mutate(lfspan=year_of_death-year_of_birth) %>% 
  filter(lfspan>=4,  lfspan<=101) %>%
  mutate(tnt=if_else(title!="", "1", "0")) %>% 
  ggplot(mapping = aes(x=tnt, y=lfspan))+
  geom_boxplot()+
  labs(title = "Distribution of lifespan for Swiss politicians",
       subtitle= "with a title and without a title",
       x="politicians without a title (0) and politicians with a title (1)",
       y="Lifespan in years")
  
  
```

At first glance there seems to be a difference in the median value of the two lifespan distributions but none of the medians falls outisde the interquartile range of the other distributions. Thus the difference may not be statistically significant. Also politicins without a title seem to have a longer spread of lifespan as the no-title group is almost twice as large as the title group (356 as compared to 195).

Now a t-test of significance.

```{r}
persons_tbl %>% 
  filter(year_of_death!="") %>%
  mutate(year_of_death=as.numeric(year_of_death)) %>% 
  filter(year_of_birth!=0) %>% 
  mutate(lfspan=year_of_death-year_of_birth) %>% 
  filter(lfspan>=4,  lfspan<=101) %>%
  mutate(tnt=if_else(title!="", "1", "0")) %>%
  mutate(tnt=factor(tnt, levels=c("0", "1"))) %>% 
  t_test(lfspan ~ tnt, order = c("0", "1"), var.equal=FALSE)

```

The t-test confirm that the difference between the two groups is statistically significant as the p-value is very small (significant at 0.05), zero is not contained in the confidence interval. It seems that people without a title lived on average 5 years less than those with a title.

## Part 6

Create a new variable which splits the politicians into 2 subgroups: one subgroup of politicians who were BORN_BEFORE_1918 (i.e. for which YEAR_OF_BIRTH < 1918), and another subgroup of politicians who were BORN_AFTER_1918 (i.e. for which YEAR_OF_BIRTH >= 1918). Reply to the questions asked in the previous part (i.e. graphical comparison and statistical test) for each of the 2 subgroups.

```{r}
persons_tbl %>% 
  filter(year_of_death!="") %>%
  mutate(year_of_death=as.numeric(year_of_death)) %>% 
  filter(year_of_birth!=0) %>% 
  mutate(lfspan=year_of_death-year_of_birth) %>% 
  filter(lfspan>=4,  lfspan<=101) %>%
  mutate(y_1918=if_else(year_of_birth>=1918, "1", "0")) %>% 
  ggplot(mapping = aes(x=y_1918, y=lfspan))+
  geom_boxplot()+
  labs(title = "Distribution of lifespan for Swiss politicians",
       subtitle= "born before or after 1918",
       x="politicians born before 1918 (0) and politicians born in and after 1918 (1)",
       y="Lifespan in years")
  
```

The boxplot suggests that there is a difference between the before 1918 group and after the 1918 group. The median for the after group also falls outside the interquartile range of the before group. Swiss politicians seem to have lived significantly longer after 1918. The before group also has a wider spread of lifespan.

Now the t-test to check for statistical significance.

```{r}
persons_tbl %>% 
  filter(year_of_death!="") %>%
  mutate(year_of_death=as.numeric(year_of_death)) %>% 
  filter(year_of_birth!=0) %>% 
  mutate(lfspan=year_of_death-year_of_birth) %>% 
  filter(lfspan>=4,  lfspan<=101) %>%
  mutate(y_1918=if_else(year_of_birth>=1918, "1", "0")) %>%
  mutate(y_1918=factor(y_1918, levels=c("0", "1"))) %>% 
  t_test(lfspan ~ y_1918, order = c("0", "1"), var.equal=FALSE)
```

The assumption is confirmed by the t-test. As the p-value is very low (significant at 0.05) and 0 is not included in the confidence interval we can reject the null hypothesis that there is no differece in means between the two groups. After 1918 Swiss politicians lived on average a little over 6 years longer.

Having discovered this new source of difference in the dataset, I wonder if title still make a difference for the lifespan of the post 1918 generation.

```{r}
persons_tbl %>% 
  filter(year_of_death!="") %>%
  mutate(year_of_death=as.numeric(year_of_death)) %>% 
  filter(year_of_birth!=0) %>% 
  mutate(lfspan=year_of_death-year_of_birth) %>% 
  filter(lfspan>=4,  lfspan<=101) %>%
  mutate(y_1918=if_else(year_of_birth>=1918, "1", "0")) %>%
  filter(y_1918=="1") %>% 
  mutate(tnt=if_else(title!="", "1", "0")) %>% 
  ggplot(mapping = aes(x=tnt, y=lfspan))+
  geom_boxplot()+
  labs(title = "Distribution of lifespan for Swiss politicians born after 1918",
       subtitle= "with a title and without a title",
       x="politicians without a title (0) and politicians with a title (1)",
       y="Lifespan in years")
```

The boxplot suggests there may not be a significant difference anymore between politicians with a title and those without.

So I'll do the t-test to check.

```{r}
persons_tbl %>% 
  filter(year_of_death!="") %>%
  mutate(year_of_death=as.numeric(year_of_death)) %>% 
  filter(year_of_birth!=0) %>% 
  mutate(lfspan=year_of_death-year_of_birth) %>% 
  filter(lfspan>=4,  lfspan<=101) %>%
  mutate(y_1918=if_else(year_of_birth>=1918, "1", "0")) %>%
  filter(y_1918=="1") %>% 
  mutate(tnt=if_else(title!="", "1", "0")) %>%
  mutate(tnt=factor(tnt, levels=c("0", "1"))) %>% 
  t_test(lfspan ~ tnt, order = c("0", "1"), var.equal=FALSE)
```

The t-test confirms (p-value above 0.05 and 0 included in confidence interval) that we can no longer reject the null hypothesis that there is no difference in the average life span of politicians with a title as compared to those without a title if looking only at those born after 1918. The difference recorded before under question 5 was most likely due to differences between the pre-1918 and post-1918 groups.




## Part 7

Which politicians have had the most mandates? Create a top 10 horizontal bar chart.

```{r}

mandates_tbl %>% 
  left_join(persons_tbl, by=c("person_id")) %>% 
  select(person_id, assembly, mandate_start_year, mandate_end_year, lastname, firstname, gender) %>% 
  mutate(lab=str_c(firstname, lastname, sep = " ")) %>%
  mutate(person_id=as.character(person_id)) %>%
  count(person_id, lab, name="no_mandates") %>% 
  arrange(desc(no_mandates)) %>% 
  top_n(10) %>%
  arrange(no_mandates)%>% 
  ggplot(mapping = aes(x=lab, y=no_mandates, label=no_mandates))+
  geom_bar(stat="identity", fill="light blue")+
  geom_text(size = 3, position = position_stack(vjust = 0.5))+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+
  labs(title = "Top 10 of Swiss politicians by total number of mandates (1804-2019)", 
       x="", 
       y="number of mandates")

```

## Part 8

Do some politicians have multiple mandates at the same time? You will need to find a way to check if some mandates started before the previous one of the same politician ended.

My strategy: use lubridate to convert start of mandate data into a date. Same for end of the mandate. Then create new variable equal to the end date of the previous mandate using lag) and filter the mandates where tha start date is smaller than the previus end date.

The answer is yes, some politicians have multiple mandates at the same time (33 distinct person_id's).

```{r}
mandates_tbl %>% 
  summarise(count=sum(is.na(mandate_start_month)))
# Checked that no missing data

id_more_mandates <- mandates_tbl %>% 
  count(person_id) %>% 
  filter(n >1) %>% 
  pull(person_id) # these are the politicians with more than one mandate, but are these mandates overlapping?

mandates_tbl %>% 
  unite(mandate_start_date, mandate_start_day, mandate_start_month, mandate_start_year, sep = "-", remove = FALSE) %>% 
  mutate(mandate_start_date= parse_date_time(mandate_start_date, orders="dmy")) %>% 
  unite(mandate_end_date, mandate_end_day, mandate_end_month, mandate_end_year, sep = "-", remove = FALSE) %>% 
  mutate(mandate_end_date= parse_date_time(mandate_end_date, orders="dmy")) %>% 
  select(id, person_id, assembly, mandate_start_date, mandate_end_date) %>%
  filter(person_id %in% id_more_mandates) %>% 
  group_by(person_id) %>% 
  arrange(person_id, mandate_start_date) %>% 
  mutate(end_date_prev=lag(mandate_end_date)) %>% 
  filter(mandate_start_date< end_date_prev) 

```

## Part 9

Have some politicians been affiliated to different parties over the years?

The short answer is yes. Below is a list of 449 person id's for which more than one party was listed. In some cases it can be that parties were entered with a similar but not identical name. In some cases the party affiliation is missing and is counted as different. 

With a treatment of missing party affiliation data we are down to 340 person id's which were affiliated to more than one party.

```{r}
affiliations_tbl %>% 
  distinct(person_id, party) %>% 
  group_by(person_id) %>% 
  summarise(no_parties=n()) %>% 
  arrange(desc(no_parties)) %>% 
  filter(no_parties>1)

# with treatment of missing data under party affiliation

affiliations_tbl %>% 
  filter(party!="") %>% 
  distinct(person_id, party) %>% 
  group_by(person_id) %>% 
  summarise(no_parties=n()) %>% 
  arrange(desc(no_parties)) %>% 
  filter(no_parties>1)

```

## Part 10

Take a sample of 20 politicians with a listed address and plot them on a {leaflet} map. You will need to use an API that converts the addresses to geocoordinates.

```{r}
# Some data cleaning, removing missing data
addresses_tbl<- 
  addresses_tbl %>% 
  clean_names() %>% 
  filter(street!="", house_number!="", postal_code!="", city!="")

# Getting a dataset and sample of 20 addresses separated by the + sign  
addy_tbl <- addresses_tbl %>% 
  select(person_id, street, house_number, postal_code, city) %>% 
  left_join(persons_tbl) %>% 
  select(-title, -year_of_birth, -year_of_death, -gender, -activity) %>% 
  sample_n(20) %>% 
  mutate(addy=str_c(house_number, street, postal_code, city, sep = "+")) %>%   select(person_id, firstname, lastname, addy)

addy_str<- addy_tbl %>% 
  pull(addy)

# Using API geocode to get latitude and longitude.

longt <- str_glue("https://geocode.xyz/{addy_str}?json=1") %>% 
  map(GET) %>% 
  keep(~status_code(.)==200) %>% 
  map(httr::content, as="parsed") %>% 
  map(magrittr::extract, c("longt", "latt"))

# As I got blocked two days ago, I'm doing this in two steps now
longlat_tib<- longt %>% 
  map(flatten_df) %>% 
  bind_rows() %>% 
  mutate(longt=as.numeric(longt)) %>% 
  mutate(latt=as.numeric(latt))

# Now binding the 20 politicians data tibble with the longitude and latitude data tibble

addy_tbl %>% 
  mutate(fullname=str_c(firstname, lastname, sep = " ")) %>% 
  bind_cols(longlat_tib) %>% 
  leaflet() %>% 
  addTiles() %>% 
  addMarkers(lng = ~ longt, lat = ~ latt, label = ~fullname, popup = ~addy)

```

