---
title: "Difference in difference with LISS propensity to vote and political orientation and IV on Ukraine war"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Read the data

We previously cleaned and joined the data in a csv balanced panel file. Here we simply read in that data.


```{r message=FALSE, warning=FALSE}
library(readr)
library(bacondecomp)
library(did)
library(eatATA)
library(haven)
library(lfe)
library(panelView)
library(tidyverse)

panel19_23_stacked <- read_csv("panel19_23_stacked.csv")
```

Balanced panel, 5 years of data.

```{r}
panel19_23_stacked %>% 
  dplyr:: group_by(survey_year) %>% 
  dplyr:: summarise(count=n())
```

Every year about 866 individuals from our panel are left-leaning and 792 are right-leaning.


```{r}
panel19_23_stacked %>% 
  dplyr:: group_by(right) %>% 
  dplyr:: summarise(count=n()/5)
```

## Create treatment variable

In our case, we need to create a treatment variable - **treatment_z**- which takes value 1 when year=2022 or 2023 and the individual is left-leaning (i.e. right==0).

Another possibility is that the effect of the crisis had not yet take center stage in 2022 at the time when the LISS data was collected (Dec 2021-Mar 2022), with the war commencing enf of Feb 2022. Thus another instrument we could try is treatment instrument **treatment_23** which takes value 1 when year=2023 (not 2022) and the individual is left-leaning.

In addition, political orientation is encoded on a 0 to 10 scale. We rescale it to 1 to 11 (to avoid having zeros).


```{r}

panel19_23_stacked <- panel19_23_stacked %>% 
  mutate(treatment_z = if_else((right==0 & survey_year>=2022), true= 1, false=0)) %>% 
  mutate(treatment_23 =if_else((right==0 & survey_year==2023), true= 1, false=0)) %>% 
  mutate(political_orientation=political_orientation+1)
```



```{r}
panel19_23_stacked %>% 
  head()
```

## Plotting the data

```{r}
panel19_23_stacked %>% 
  group_by(right, survey_year) %>% 
  summarise(mean_prop_to_vote= (mean(propensity_to_vote))) %>% 
  mutate(right=as.factor(right)) %>% 
  ggplot(aes(x=survey_year, y=mean_prop_to_vote, color=right))+
  geom_point()+
  geom_line()+
  labs(title="Mean propensity to vote by political orientation",
       subtitle = "1=right-leaning, 0=left-leaning, no swing",
       caption="LISS Panel data 2019-2023: 1658 observations")
```


```{r}
panel19_23_stacked %>% 
  mutate(right=as.factor(right)) %>% 
  ggplot(aes(x=propensity_to_vote, y=right, color=right)) +
  geom_point()+
  geom_boxplot() +
  coord_flip() +
  facet_wrap(vars(survey_year)) +
  labs(title="Distribution of propensity to vote by political orientation",
       x="propensity to vote (0 to 100)",
       y="")
```


## TSLS with IV=time of the Russia-Ukraine war for the left-leaning group

Now we estimate the first stage of our dif-in-dif model

```{r}
dd_reg <- felm(political_orientation ~ treatment_z | survey_year + nomem_encr, data=panel19_23_stacked)

summary(dd_reg)
```


We can see that it seems that the treatment coefficient is highly significant. We'd have to run a parallel trends regression to see if the parallel trends assumption holds and the estimators on lagged treatment are within the confidence interval.

But for now let's store the main_ATT_1 and the confidence interval boundaries.

```{r}
main_ATT_1= summary(dd_reg)$coefficients["treatment_z", "Estimate"]

ATT_1_CI_lower_boundary = abs(main_ATT_1 - qnorm(0.975)*summary(dd_reg)$coefficients["treatment_z", "Std. Error"])

ATT_1_CI_upper_boundary = abs(main_ATT_1 + qnorm(0.975)*summary(dd_reg)$coefficients["treatment_z", "Std. Error"])
```


```{r}
c(main_ATT_1, ATT_1_CI_lower_boundary, ATT_1_CI_upper_boundary)
```


For the second stage of our two-stage least-squares (TSLS) we need to store the predicted values of political orientation from the first stage model above (dd_reg). We'll store them in a new variable called pred_political_orientation.



```{r}
panel19_23_stacked <- panel19_23_stacked %>% 
  mutate(pred_political_orientation = as.tibble(fitted(dd_reg)) %>% 
  pull(political_orientation))

```


Just to get an idea of how predicted political_orientation and true political orientation compare, we'll make a scatterplot.

```{r}
panel19_23_stacked %>% 
  mutate(right=as.factor(right)) %>% 
  ggplot(aes(x=political_orientation, y=pred_political_orientation, color=right)) +
  geom_point() +
  facet_wrap(vars(survey_year)) +
  labs(title="Comparison between political orientation and predicted political orientation")
```


For the second stage, we regress the outcome (propensity to vote) on the predicted political orientation, with fixed effects.


```{r}
dd_reg_2 <- felm(propensity_to_vote ~ pred_political_orientation | survey_year + nomem_encr, data=panel19_23_stacked)

summary(dd_reg_2)
```


Our LATE estimand is highly insignificant. We store the estimated values and confidence intervals.


```{r}
LATE= summary(dd_reg_2)$coefficients["pred_political_orientation", "Estimate"]

LATE_CI_lower_boundary = abs(LATE - qnorm(0.975)*summary(dd_reg_2)$coefficients["pred_political_orientation", "Std. Error"])

LATE_CI_upper_boundary = abs(LATE + qnorm(0.975)*summary(dd_reg_2)$coefficients["pred_political_orientation", "Std. Error"])

c(LATE, LATE_CI_lower_boundary, LATE_CI_upper_boundary)

```


## Standard OLS estimation 

We regress propensity to vote on political orientation, the treatment variable (=1 after the crisis, for left-leaning people) and the interaction term between treatment and political orientation.

Probably the relationship is endogenous, but just to be able to compare results.

```{r}
library(fixest)

ols_reg <- feols(
  propensity_to_vote ~ political_orientation, # Regression formula
  data=panel19_23_stacked,
  vcov = "hc1" #--variance-covariance ratio, hc=heterochedasticity-consistent
)

summary(ols_reg)
```

Here we obtain a negative and significant coefficient. Increasing political orientation by one unit = moving to the right is associated with 0.356 lower propensity to vote.



```{r}
library(binsreg)

binscatter <- binsreg(panel19_23_stacked$propensity_to_vote, panel19_23_stacked$political_orientation)
binscatter$bins_plot +
  labs(y = "Propensity to vote", x = "Political orientation")
```


Simple TSLS, no fixed, effects:

```{r}
feols(
  propensity_to_vote ~ 1 | 0 | political_orientation ~ treatment_z, 
  data = panel19_23_stacked,
  vcov = "hc1"
)
```

TSLS with fixed effects. Here we obtain the exact same results as above when implementing TSLS in two stages.


```{r}
feols(
  propensity_to_vote ~ 1 | survey_year + nomem_encr | political_orientation ~ treatment_z, 
  data = panel19_23_stacked,
  vcov = "hc1"
)
```


And a Hasman test, which is highly significant, so we shoould not use OLS results.

```{r}
library(ivDiag)

eff_F(panel19_23_stacked, Y = "propensity_to_vote", D = "political_orientation", Z = "treatment_z")
```



## Repeating TSLS for treatment_23 (if we assume the crisis is only apparent in the 2023 data)


Another possibility is that the effect of the crisis had not yet take center stage in 2022 at the time when the LISS data was collected (Dec 2021-Mar 2022), with the war commencing enf of Feb 2022. Thus another instrument we could try is treatment instrument **treatment_23** which takes value 1 when year=2023 (not 2022) and the individual is left-leaning.


```{r}
feols(
  propensity_to_vote ~ 1 | survey_year + nomem_encr | political_orientation ~ treatment_23, 
  data = panel19_23_stacked,
  vcov = "hc1"
)
```


The coefficient estimate increases and the p-value improves, but the conclusion does not change. We cannot reject the null hypothesis that there is no difference in the impact the Ukraine-Russian was had on left-leaning as compared to right-leaning Dutch voters.

And a Hausman test, which still is highly significant, so we should not use the OLS results, which are also reported below for completeness.


```{r}
eff_F(panel19_23_stacked, Y = "propensity_to_vote", D = "political_orientation", Z = "treatment_23")
```



