---
title: "R Notebook"
output: github_document
---
# FDS Final Project: Report #4

This project looks at a dataset about athletes at the Olympic Games in 4 different sheets.
I first import the data with the chunk below.

```{r warning=FALSE}
library(readxl)
athletes <- read_excel("../Report_4/olympics.xlsx", 
    sheet = "athletes")

country <- read_excel("../Report_4/olympics.xlsx", 
    sheet = "country")

games <- read_excel("../Report_4/olympics.xlsx", 
    sheet = "games")

medals <- read_excel("../Report_4/olympics.xlsx", 
    sheet = "medals")

```

As this chunck was quite long I load any additional libraries with the chunck below. These are R packages that I need to use in the rest of the project.

```{r message=FALSE, warning=FALSE}
library(janitor)
library(tibble)
library(dplyr)
library(tidyr)
library(visdat)
library(ggplot2)
library(stringr)
library(infer)
library(knitr)
```

Next I will do a bit of data cleaning, including cleaning column names, and renaming a few to avoid any further confusion. There will be more data cleaning when needed later on.

```{r}
games <-clean_names(games) %>% 
  rename(gm=games)
medals<-clean_names(medals) %>% 
  rename(id=athlete_id) %>% 
  rename(gm=games)
country<-clean_names(country) %>% 
  rename(id=athlete_id) %>% 
  rename(gm=games)
athletes<-clean_names(athletes)

```

Before getting started, I check what happened at the 1956 games in the games data file. There are two different entries for the summer Olympics, which were party held in Melbourne and party in Stockholm.

```{r}
games %>% 
  filter(year=="1956")
```
This means I need to be careful not to lose this data when combining the worksheets together. Wikipedia suggests that the 1956 Summer Olympics was held in Melbourne, with the exception of the equestrian events, which were held in Stockholm. For the purpose of the analysis I'll replace Stockholm with Melbourne for the year 1956 to avoid the double counting.

```{r}
games<- games %>% 
  filter(city!="Stockholm" | year!= 1956)

```



## Part 1. Have some athletes competed for different countries over time?

The answer is yes. I first combine the country and games worksheet to have access to "noc" =country and "year" in the same dataset. Then I create a vector of athlete id values (id) for which more than one countries are listed over the years. For example id 116 competed for both UAR and EGY. Some competed for as many as 4 different country codes - as was the case for id 66953. The "more than one country id's are stored in the vector id_more_than_1c.

```{r}

oly1<- left_join(country, games) %>% 
  select(-gm)

id_more_than_1c <- oly1 %>% 
  distinct(id, noc) %>% 
  group_by(id) %>% 
  count(name="nr_delegations") %>% 
  filter(nr_delegations!=1) %>% 
  arrange(desc(nr_delegations)) %>% 
  pull(id)

oly1 %>% 
  filter(id %in% id_more_than_1c)
```

## Part 2. Who are the ten athletes that took part in most games?

I first combine the athletes dataset with the previous country and games (oly1) tibble. In this new "oly 2" dataset each athlete appears the number of times he or she has competed, just like in "country" (in athletes each id appeared only once). Then I create a tibble with the list of top 10 names and the number of games in which they participated.


```{r}
oly2<-left_join(athletes, oly1)

oly2 %>% 
  count(name) %>% 
  arrange(desc(n)) %>% 
  top_n(10)

```
The list of athletes that took part in most games includes Ian Millar with 10 games, then Afanasijs Kuzmins and Hubert Raudaschl each with 9 games, followed by 12 other athletes with 8 games each. Thus the top 10 should include 15 athletes.


## Part 3. What athlete(s) kept a Gold medal for the longest time?

For this part I need to combine together the medal, athletes and games tibbles, but with certain other considerations. 

I first do some more data cleaning on the "medals" dataset. It seems that some atheletes are listen more than once in this dataset, especially for unknown art events. I look for distinct/unique combinations of id, game, event and medal.

Then I tried to join oly2 with medals but this generated some further double counting in the data and the total number of Gold medals increased. 

So I changed my strategy and joined medals with athetes and then with games to create "oly_new", which worked better for my purpose.

One more clarification, keeping a Gold medal means I need to look for consecutive years when someone had a Gold medal.



```{r}
# To show "medals" dataset has repeat entries of atheletes per game and event.
medals %>% 
  count(id, gm, event, name="nr_games") %>% 
  arrange(desc(nr_games))

# An example of repeat entries.
medals %>% 
  filter(id==83312)

medals <-medals %>% 
  distinct(id, gm, event, sport, medal)

oly_new<- left_join(medals, athletes) %>% 
  left_join(games)

q3 <- oly_new %>% 
  filter(medal=="Gold") %>% 
  group_by(id, event) %>% 
  mutate(prevyear=lag(year, default=0)) %>% 
  mutate(diffyear=year-prevyear) %>% 
  mutate(diff4=if_else(diffyear==4, 1, 0)) %>% 
  mutate(diffsum=cumsum(diff4)) %>% 
  select(-gm, -medal, -sport, -height, -weight, -sex, -city) %>% 
  arrange(desc(diffsum)) %>% 
  ungroup() %>% 
  top_n(1, diffsum) %>% 
  pull(name)


```



The answer is `r q3`. He was the only one to keep his Gold medal for fencing for 4 consecutive years.


## Part 4. What country(ies) kept a Gold medal for the longest time?

To do this I finally join together all datasets, in the order of medals with athletes, with games and with country into an oly_all dataset, which has no duplication problems (I checked).

The question refers to the country with the largest number of gold medals for the same event per year (considering different athletes over time), in consecutive competitions. 

I had to create a function ff that resets cumsum() every time the value 0 is encoutered. This is because in the dataset there are series of consecutive gold medals but with breaks in between. This ws not the case for individual atheletes.

Kenya and the United States both share the record for keeping the gold medal for 8 consecutive Olymping games ending in 2016. Kenya for Athletics Men's 3,000 metres Steeplechase and the United States for Swimming Men's 4 x 100 metres Medley Relay.

```{r}
oly_all<- left_join(medals, athletes) %>% 
  left_join(games) %>% 
  left_join(country)

ff = function(x)
{
    cs = cumsum(x)
    cs - cummax((x == 0) * cs)
}

oly_all %>% 
  select(noc, event, id, year, medal) %>% 
  filter(medal=="Gold") %>% 
  distinct(noc, event, year) %>%
  arrange(noc, event, year) %>% 
  group_by(noc, event) %>% 
  mutate(prevyear=lag(year, default = 0)) %>% 
  mutate(diffyear=year-prevyear) %>% 
  mutate(diff4=if_else(diffyear==4, 1, 0)) %>%
  mutate(diffsum=ff(diff4)) %>% 
  arrange(desc(diffsum)) %>%  
  ungroup() %>% 
  top_n(1)


```

Just checking to see what the data looks like for one of the countries with the largest number of gold medals by event.

```{r}
oly_all %>% 
  filter(medal=="Gold", noc=="KEN", event=="Athletics Men's 3,000 metres Steeplechase") %>% 
  arrange(year)
```


## Part 5. Who are the ten athletes that competed in the most events (some athletes take part in more than one event during games) ?

I first discovered that in the atheletes file there are a few athletes that have the same name but different height, weight and id. For example Ivan Ivanov. Thus the name is not a unique identifyer in this dataset, so I use "id" instead.

The top 10 athletes who competed in most events are listed below, starting with the gymnast Heikki Ilmari Savolainen of Finland who competed 39 times over 5 years. The list of top 10 includes 12 as the last three althletes listed all competed 30 times.

```{r}
athletes %>% 
  count(name, sex, name="no_ids") %>% 
  arrange(desc(no_ids))

# Showing one example where name is not a unique identifier.
athletes %>% 
  filter(name=="Ivan Ivanov")

oly_all %>% 
  distinct(id, name, event, year, season) %>% 
  count(id, name, name="no_games") %>% 
  arrange(desc(no_games)) %>% 
  top_n(10)

```

Just checking to see what the data looks like for the first one in the list.

```{r}
oly_all %>% 
  filter(name=="Heikki Ilmari Savolainen")
```


## Part 6. Create a new table showing the number of medals per country (rows) and per year (column). Keep only the 15 countries with the most medals overall.

I first create a tibble in the exsiting format including the number of medals (any) by country(noc), year and create a new variable for the sum of all medals over all year (y_all). Then I sort the data by year. Next I create a wider pivot with the years as columns, as well as my y_all total medals variable used to filter only the top 15 countries. The first three countries are USA, URS and GER.

Note to self. In the code below I added a parameter in the pivot_wider()function to set missing values to 0 (i.e. no medal).

```{r}

medal_tib<- oly_all %>% 
  filter(!is.na(medal)) %>% 
  count(noc, year, name="freq") %>% 
  group_by(noc) %>% 
  mutate(y_all=(sum(freq))) %>% 
  arrange(year) %>% 
  pivot_wider(names_from = "year", values_from = "freq", values_fill = list(freq=0)) %>% 
  arrange(desc(y_all)) %>% 
  head(15) %>% 
  ungroup()

medal_tib %>% 
  select(noc, y_all, "2000", "2002", "2004", "2006", "2008", "2010", "2012", "2014", "2016") %>% 
  kable()

```

## Part 7. Is there a relationship between country and the probability of winning a medal?

We know that some countries win more medals than others. In this part we want to work out whether some countries win more medals simply because they have more athletes competing for more events, or whether their athletes are in fact performing better in general.

For this part, look only at the top 15 countries that you established in part 6.

First, create a horizontal barchart listing for each of the fifteen countries the percentage of medals won out of all medals competed for by that country.

```{r}
medal_tib <- medal_tib %>% 
  select(noc, y_all) %>% 
  arrange(noc)

countr_15 <- medal_tib %>% 
  pull(noc)

competed_tib<- oly_all %>% 
  filter(noc %in% countr_15) %>% 
  count(noc, year, name="competed") %>% 
  group_by(noc) %>% 
  mutate(y_comp=(sum(competed))) %>% 
  pivot_wider(names_from = "year", values_from = "competed") %>% 
  select(noc, y_comp) %>% 
  arrange(noc) %>% 
  bind_cols(medal_tib) %>% 
  mutate(success=y_all/y_comp*100)

competed_tib %>% 
  ggplot(mapping=aes(x=noc, y=success, label=success))+
  geom_bar(stat="identity", fill="light blue")+
  geom_text(aes(label = round(success, digits = 0)), size = 3, position = position_stack(vjust = 0.95), check_overlap = TRUE)+
  labs(title= "Percentage of medals won out of medals competed",
       subtitle = "for the top 15 countries by total number of medals (1896 to 2016)",
       x= "country",
       y= "per cent")+
  theme_classic()

```

Run a Chi Square test to test the hypothesis: "There is no relationship between country and whether they win a medal or not".

To do tis I first transform the column medal into a factor variable that takes on 0 for no medal and 1 for when a medal was won.

```{r}
oly_all<- oly_all %>% 
  mutate(medal=replace(medal, !is.na(medal), 1)) %>% 
  mutate(medal=replace(medal, is.na(medal), 0)) %>% 
  mutate(medal=factor(medal, levels = c(0,1)))

# A contingency table
oly_all %>% 
  count(noc, medal, name="freq")

chisq_test(oly_all, formula = noc ~ medal)

```

The result of the chi-squared test is highly significant as the p-value is very low meaning that there is a significant relationship between country and the possibility of winning a medal.

## Part 8

Create a scatterplot showing the average height and weight of competitors per sport (one dot per sport). Add labels with the sport names for:

    the largest average height
    the largest average weight
    the smallest average height
    the smallest average weight
    the largest average BMI
    the smallest average BMI

It might be that the same dot qualifies for multiple labels. 

In the first part I deal with the missing data, I create the bmi variable and join together the data.
In the second part I create a new tibble "variable" with the list of values to annotate on the chart.
Then I create a new variable "annot" inside the original dataset with the labels needed for the data to annotate.
I notice that maxweight coincides with maxbmi and minweight coincides with minbmi. I correct the "annot" code for this.

```{r}
#I'll first have a look at how much data is missing
oly_all %>% 
  summarise(count=sum(is.na(height)))

oly_all %>% 
  summarise(count=sum(is.na(weight)))

aveweight<- oly_all %>% 
  group_by(sport) %>% 
  summarise(ave_height=mean(height/100, na.rm = TRUE),
            ave_weight=mean(weight, na.rm = TRUE)) %>% 
  arrange(sport) %>% 
  mutate(ave_height=replace(ave_height, is.na(ave_height), median(ave_height, na.rm=TRUE)), 
         ave_weight=replace(ave_weight, is.na(ave_weight), median(ave_weight, na.rm=TRUE)),
         bmi=ave_weight/ave_height^2)

values <- aveweight %>% 
  summarise(maxweight=max(ave_weight),
            minweight=min(ave_weight),
            maxheight=max(ave_height),
            minheight=min(ave_height),
            maxbmi=max(bmi),
            minbmi=min(bmi))

annot_aveweight<- aveweight %>% 
  mutate(annot=case_when(ave_weight==max(ave_weight)~"max weight & max bmi",
                         ave_weight==min(ave_weight)~"min weight & min bmi",
                         ave_height==max(ave_height)~"max height",
                         bmi==max(bmi)~"max bmi",
                         bmi==min(bmi)~"min bmi",
                         ave_height==min(ave_height)~"min height")) %>% 
  mutate(annot=replace(annot, is.na(annot), ""))
```

I had previously written "annot" with if_else() because I though case_when() only takes conditions on one variable (weight, height or bmi), but I was wrong.

```{r}

# annot_aveweight <- aveweight %>% 
  #mutate(annot=if_else(
    #ave_weight==max(ave_weight, na.rm=TRUE), "maxweight", 
    #if_else(
      #ave_weight==min(ave_weight, na.rm=TRUE), "minweight", 
      #if_else(
        #ave_height==max(ave_height, na.rm=TRUE), "maxheight",
        #if_else(
          #ave_height==min(ave_height, na.rm=TRUE), "minheight",
          #if_else(
            #bmi == max(bmi, na.rm=TRUE), "maxbmi",
            #if_else(
              #bmi == min(bmi, na.rm=TRUE), "minbmi", "0")))))))

```



Below is the code for the scatter plot.

```{r}
annot_aveweight %>% 
  ggplot(mapping = aes(x=ave_height, y=ave_weight))+
  geom_point(mapping = aes(color=annot, na.rm=TRUE), shape=16, size=3)+
  geom_text(aes(label=annot, na.rm=TRUE))+
  labs(title ="Average height vs average weight per sport at the Olympics",
       x="average height (m)",
       y="average weight (kg)")+
  theme(legend.position = "none")+
  scale_x_continuous(limits = c(1.6, 1.95))

```
## Part 9

Create a line plot showing the number of medals given by year (one line for Gold, one line for Silver and one line for Bronze). Does it change over time? Use facet to separate the medals at Summer games and Winter games.

```{r}

oly_new %>% 
  mutate(medal=replace(medal, is.na(medal), "")) %>% 
  filter(medal!="") %>%
  mutate(medal=factor(medal, labels=c("Gold", "Silver", "Bronze"))) %>%
  count(medal, year, season, name="freq") %>%   
  ggplot(mapping = aes(x=year, y=freq, color=medal))+
  geom_line(size=1)+
  facet_wrap(vars(season), ncol=1)+
  labs(title="Number of total Olympic medals awarded by year",
       y="number")+
  theme_light()

```


