---
title: "FDS Final Project: Report #3"
output: github_document
---

## Part 1

In this project, I use an API to get your data from the  "The Movie DB", https://www.themoviedb.org/.

```{r message=FALSE, warning=FALSE}
library(httr)
library(purrr)
library(magrittr)
library(tibble)
library(dplyr)
library(stringr)
library(tidyr)
library(ggplot2)
library(infer)
```


Try to send a GET request to some of the example queries and inspect the result. 

Below is the request for the movie with id 550.

```{r warning=FALSE}
mov_550 <- GET("https://api.themoviedb.org/3/movie/550?api_key=32dcb877caa66aa7c0cf44a3fbf215ce")

content(mov_550) %>% 
pluck("original_title")

```


Inspired just by these examples, how would create new requests:
What are the highest grossing dramas from 2010? See below

```{r}
mov_gross_2010<- GET("https://api.themoviedb.org/3/discover/movie?api_key=32dcb877caa66aa7c0cf44a3fbf215ce&certification_country=US&certification=R&sort_by=revenue.desc&primary_release_year=2010&with_genres=18")

content(mov_gross_2010, as="parsed") %>% 
  pluck("results") %>% 
  map(pluck, "original_title")
```


Have Will Ferrell and Liam Neeson even been in a movie together?
The answer is yes, in 4 movies.

```{r}
mov_2ppl<- GET("https://api.themoviedb.org/3/discover/movie?api_key=32dcb877caa66aa7c0cf44a3fbf215ce&with_people=23659,3896")

content(mov_2ppl, as="parsed") %>% 
  pluck("results") %>% 
  map(pluck, "original_title")
```


Can you find kids movies with Tom Cruise in them? Yes

```{r}
mov_kidTC<- GET("https://api.themoviedb.org/3/discover/movie?api_key=32dcb877caa66aa7c0cf44a3fbf215ce&certification_country=US&certification.lte=G&with_cast=500&sort_by=popularity.desc")

content(mov_kidTC, as="parsed") %>% 
  pluck("results") %>% 
  map(pluck, "original_title")
```

## Part 2

From RStudio, what query would you make to find the id of the animation movies company "Pixar"? Show both the query and how you extract the id from the result in your report.

I used the get/search/company endpoint. There are two id's stored for Pixar: 3 and 125928. As the company stored under 3 has full details (such as origin country)I think this must be the correct id for Pixar.

```{r}
mov_pix <- GET("https://api.themoviedb.org/3/search/company?api_key=32dcb877caa66aa7c0cf44a3fbf215ce&query=Pixar&page=1")

content(mov_pix, as="parsed") %>% 
  pluck("results") %>% 
  map(pluck, "id")
```

## Part 3

Now that we have the id of Pixar, we should be able to find all the movies that they have worked on. Write a query that will give you all the Pixar movies and sort them by descending revenue. The result will be given to you as a JSON (parsed to a list by {httr}). 

Convert this list to a tibble so you have one row per film and one column per interesting piece of information. If you want to get the other pages, you need to play with the page parameter in your url.

I used a GET/discover/movie/with_companies request and I saw that there are six pages of results. Below I create a request for each of the 6 pages.

```{r}
mov_pix1<- GET("https://api.themoviedb.org/3/discover/movie?api_key=32dcb877caa66aa7c0cf44a3fbf215ce&with_companies=3&sort_by=revenue.desc&page=1")

mov_pix2<- GET("https://api.themoviedb.org/3/discover/movie?api_key=32dcb877caa66aa7c0cf44a3fbf215ce&with_companies=3&sort_by=revenue.desc&page=2")

mov_pix3<- GET("https://api.themoviedb.org/3/discover/movie?api_key=32dcb877caa66aa7c0cf44a3fbf215ce&with_companies=3&sort_by=revenue.desc&page=3")

mov_pix4<- GET("https://api.themoviedb.org/3/discover/movie?api_key=32dcb877caa66aa7c0cf44a3fbf215ce&with_companies=3&sort_by=revenue.desc&page=4")

mov_pix5<- GET("https://api.themoviedb.org/3/discover/movie?api_key=32dcb877caa66aa7c0cf44a3fbf215ce&with_companies=3&sort_by=revenue.desc&page=5")

mov_pix6<- GET("https://api.themoviedb.org/3/discover/movie?api_key=32dcb877caa66aa7c0cf44a3fbf215ce&with_companies=3&sort_by=revenue.desc&page=6")

```

Below I convert the extracted lists into tibbles, then I used bind_rows() to join them up into a single tibble with 108 rows.

```{r}

pix1<- content(mov_pix1, as="parsed") %>% 
  pluck("results") %>% 
  map(magrittr::extract, c("id", "original_title" ,"popularity", "vote_count", "vote_average", "release_date")) %>% 
  map(flatten_df) %>% 
  bind_rows()

pix2 <- content(mov_pix2, as="parsed") %>% 
  pluck("results") %>% 
  map(magrittr::extract, c("id", "original_title" ,"popularity", "vote_count", "vote_average", "release_date")) %>% 
  map(flatten_df) %>% 
  bind_rows()

pix3 <- content(mov_pix3, as="parsed") %>% 
  pluck("results") %>% 
  map(magrittr::extract, c("id", "original_title" ,"popularity", "vote_count", "vote_average", "release_date")) %>% 
  map(flatten_df) %>% 
  bind_rows()

pix4 <- content(mov_pix4, as="parsed") %>% 
  pluck("results") %>% 
  map(magrittr::extract, c("id", "original_title" ,"popularity", "vote_count", "vote_average", "release_date")) %>% 
  map(flatten_df) %>% 
  bind_rows()

pix5 <- content(mov_pix5, as="parsed") %>% 
  pluck("results") %>% 
  map(magrittr::extract, c("id", "original_title" ,"popularity", "vote_count", "vote_average", "release_date"))%>%
  map(flatten_df) %>% 
  bind_rows()

pix6 <- content(mov_pix6, as="parsed") %>% 
  pluck("results") %>% 
  map(magrittr::extract, c("id", "original_title" ,"popularity", "vote_count", "vote_average", "release_date")) %>% 
  map(flatten_df) %>% 
  bind_rows()

pix_all <- bind_rows(pix1, pix2, pix3, pix4, pix5, pix6) %>% 
  arrange(desc(vote_count)) %>% 
  mutate(company="Pixar")

glimpse(pix_all)
```

## Part 4. 

You may know that Pixar was acquired by Disney in 2006, after they had already been collaborating on films for more than a decade. For the last part of the report, we are going to look into whether this was a smart strategic decision by Disney, by comparing the popularity of both Disney and Pixar films that came out from 2006 onwards.

    *First, acquire the "id" for Disney using the search endpoint. Note that if you try to find the company ID for Disney, there will be more than one result (Disney has many subsidiaries with the name "Disney" in it). For this exercise, specifically look for "Walt Disney Pictures" in the USA.
    
```{r}
mov_disney <- GET("https://api.themoviedb.org/3/search/company?api_key=32dcb877caa66aa7c0cf44a3fbf215ce&query=Disney&page=1")

content(mov_disney, as="parsed") %>% 
  pluck("results") %>% 
  map(magrittr::extract, c("id", "name", "origin_country")) %>% 
  map(flatten_df) %>%
  bind_rows() %>% 
  filter(name=="Walt Disney Pictures")
```
    
    I find that the id for the Walt Disney Pictures is 2.
    
    
    *Second, get the vote averages and vote counts for films from Walt Disney Productions and from Pixar using the discover/movies endpoint. Use the API documentation to find out how to get films from 2006 onwards.
    
    There are 34 pages of films from Disney, so I need to either automate the data collection from pages, or limit the search to films released after 2006. I do this by using primary_release_date.gte and .lte. Now down to 8 pages. That's better :-) 
    
```{r}

mov_dis1<- GET("https://api.themoviedb.org/3/discover/movie?api_key=32dcb877caa66aa7c0cf44a3fbf215ce&with_companies=2&primary_release_date.gte=2006-01-01&primary_release_date.lte=2020-01-22&page=1")

mov_dis2<- GET("https://api.themoviedb.org/3/discover/movie?api_key=32dcb877caa66aa7c0cf44a3fbf215ce&with_companies=2&primary_release_date.gte=2006-01-01&primary_release_date.lte=2020-01-22&page=2")

mov_dis3<- GET("https://api.themoviedb.org/3/discover/movie?api_key=32dcb877caa66aa7c0cf44a3fbf215ce&with_companies=2&primary_release_date.gte=2006-01-01&primary_release_date.lte=2020-01-22&page=3")

mov_dis4<- GET("https://api.themoviedb.org/3/discover/movie?api_key=32dcb877caa66aa7c0cf44a3fbf215ce&with_companies=2&primary_release_date.gte=2006-01-01&primary_release_date.lte=2020-01-22&page=4")

mov_dis5<- GET("https://api.themoviedb.org/3/discover/movie?api_key=32dcb877caa66aa7c0cf44a3fbf215ce&with_companies=2&primary_release_date.gte=2006-01-01&primary_release_date.lte=2020-01-22&page=5")

mov_dis6<- GET("https://api.themoviedb.org/3/discover/movie?api_key=32dcb877caa66aa7c0cf44a3fbf215ce&with_companies=2&primary_release_date.gte=2006-01-01&primary_release_date.lte=2020-01-22&page=6")

mov_dis7<- GET("https://api.themoviedb.org/3/discover/movie?api_key=32dcb877caa66aa7c0cf44a3fbf215ce&with_companies=2&primary_release_date.gte=2006-01-01&primary_release_date.lte=2020-01-22&page=7")

mov_dis8<- GET("https://api.themoviedb.org/3/discover/movie?api_key=32dcb877caa66aa7c0cf44a3fbf215ce&with_companies=2&primary_release_date.gte=2006-01-01&primary_release_date.lte=2020-01-22&page=8")

```
    
Now collecting the data and compiling into tibble for Disney, like before for Pixar.

```{r}
dis1<- content(mov_dis1, as="parsed") %>% 
  pluck("results") %>% 
  map(magrittr::extract, c("id", "original_title" ,"popularity", "vote_count", "vote_average")) %>% 
  map(flatten_df) %>% 
  bind_rows()

dis2 <- content(mov_dis2, as="parsed") %>% 
  pluck("results") %>% 
  map(magrittr::extract, c("id", "original_title" ,"popularity", "vote_count", "vote_average")) %>% 
  map(flatten_df) %>% 
  bind_rows()

dis3 <- content(mov_dis3, as="parsed") %>% 
  pluck("results") %>% 
  map(magrittr::extract, c("id", "original_title" ,"popularity", "vote_count", "vote_average")) %>% 
  map(flatten_df) %>% 
  bind_rows()

dis4 <- content(mov_dis4, as="parsed") %>% 
  pluck("results") %>% 
  map(magrittr::extract, c("id", "original_title" ,"popularity", "vote_count", "vote_average")) %>% 
  map(flatten_df) %>% 
  bind_rows()

dis5 <- content(mov_dis5, as="parsed") %>% 
  pluck("results") %>% 
  map(magrittr::extract, c("id", "original_title" ,"popularity", "vote_count", "vote_average")) %>% 
  map(flatten_df) %>% 
  bind_rows()

dis6 <- content(mov_dis6, as="parsed") %>% 
  pluck("results") %>% 
  map(magrittr::extract, c("id", "original_title" ,"popularity", "vote_count", "vote_average")) %>% 
  map(flatten_df) %>% 
  bind_rows()

dis7 <- content(mov_dis7, as="parsed") %>% 
  pluck("results") %>% 
  map(magrittr::extract, c("id", "original_title" ,"popularity", "vote_count", "vote_average")) %>% 
  map(flatten_df) %>% 
  bind_rows()

dis8 <- content(mov_dis8, as="parsed") %>% 
  pluck("results") %>% 
  map(magrittr::extract, c("id", "original_title" ,"popularity", "vote_count", "vote_average")) %>% 
  map(flatten_df) %>% 
  bind_rows()

dis_all <- bind_rows(dis1, dis2, dis3, dis4, dis5, dis6, dis7, dis8) %>%  
  arrange(desc(vote_count)) %>% 
  mutate(company="Disney")

glimpse(dis_all)
```

Now combining post 2006 Pixar and Disney data together and limiting to vote_count of at least 50.

A suggestion here would be to filter the data by including only films with a vote_count of at least 50. Consider that if only a few people voted on the film, the vote average will not be as representative as when lots of people have voted on the film.

```{r}
pix_dis<- pix_all %>% 
  filter(release_date>"2005.12.31") %>% 
  select(-release_date) %>% 
  bind_rows(dis_all) %>% 
  filter(vote_count>=50)

glimpse(pix_dis)
```

*Now, compare the vote averages using boxplots and a t-test, with the aim of answering the question Are the films from Pixar on average more popular than the films from Walt Disney Pictures?

First the boxplot

```{r}

pix_dis %>% 
  ggplot(mapping = aes(x=company, y=vote_average))+
  geom_boxplot()+
  labs(title = "Distribution of average votes by company",
       subtitle= "films released in 2006 or earlier, which received at least 50 votes",
       x="company",
       y="average number of votes")
```

The boxplot suggests that indeed there is a significant difference between the two companies by average number of votes. The median from Pixar falls outside the inter-quartile range for Disney.

Now doing the t-test.

```{r}
pix_dis %>% 
  mutate(company=factor(company, levels=c("Disney", "Pixar"))) %>% 
  t_test(vote_average ~ company, order = c("Disney", "Pixar"), var.equal=FALSE)
```

The t-test confirms that Pixar movies are on average more popular than Disney films.
    
Below I automate the collection of data for all the 8 pages at the same time for Disney movies. This can be used instead of the long way in parts 3 and 4 above but I prefer to keep both versions of the code. All of the others steps are exactly the same.
    
```{r}
vec8 <- seq(from=1, to=8)

mov_dis<- str_glue("https://api.themoviedb.org/3/discover/movie?api_key=32dcb877caa66aa7c0cf44a3fbf215ce&with_companies=2&primary_release_date.gte=2006-01-01&primary_release_date.lte=2020-01-22&page={vec8}") %>% 
  map(httr::GET) %>% 
  map(httr::content, as="parsed") %>% 
  map(purrr::pluck, "results")

disney_tb<- mov_dis %>% 
  flatten() %>% 
  map(magrittr::extract, c("id", "original_title" ,"popularity", "vote_count", "vote_average")) %>% 
  map(flatten_df) %>% 
  bind_rows() %>% 
  mutate(company="Disney")

glimpse(disney_tb)

```
    
    
    


